---
title: "Master Training Data Maker"
author: "Thomas Estabrook"
date: "8/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(stringr)
library(sf)
library(cartography)
```

Goal: Get every dataset into a single, usable shapefile. This means that all trees have their proper code, invalid or dead trees are eliminated. The final file should have the following fields:
- Tree code
- DBH
- Source file
- Geometry

Outline:
1. Load in survey trees shapefile (output of survey_to_points.rmd), campus trees, arb trees, and street trees
  a. Get all trees into the correct coordinate reference system (crs)
2. Survey trees
  a. Combine with single point layer
  b. Eliminate typos
3. Campus trees, Arb trees, Street trees
  a. Recode to project tree codes
  b. Calculate proper DBH (campus)
  c. Eliminate extraneous columns
  d. Eliminate dead trees
4. Merge all into single layer

Notes
- NEED TO FILTER OVERTOPPED TREES FROM SURVEY DATA
- The table joins will produce duplicate entries UNLESS the key columns include only unique values. Unfortunately, for these datasets we will be using the `Project_Code`, `Botanical`, and `Common_Name` columns for joins, so they all need to only have unique entries. If using this script, make sure you have the most recent version of `GenusSpecies_PlantCode_TGE_edits.csv`. 
----------------------------------------------------------------------

First we load in the necessary training data layers and ensure they are all in the same projection:
```{r}
# set up tree codes table
tree_codes <- read.csv("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data//CodedTrainingData/GenusSpecies_PlantCode_TGE_edits.csv")
colnames(tree_codes)[1] <- "Common_Name"
tree_codes$Project_Code[tree_codes$Project_Code == ""] <- NA
tree_codes$Plant_Code[tree_codes$Plant_Code == ""] <- NA
tree_codes <- tree_codes %>%
  mutate(Project_Code = coalesce(Project_Code, Plant_Code)) %>%
  mutate(Common_Name = str_to_title(Common_Name))

# For the joins to work, we need to check for duplicates in the project codes. If any of these tables have an entry higher than one, you will need to adjust the plant code document accordingly. 
n_occur_code <- data.frame(table(tree_codes$Project_Code))
n_occur_common <- data.frame(table(tree_codes$Project_Code))
n_occur_botanical <- data.frame(table(tree_codes$Project_Code))

# load data layers
survey_trees_plot <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/tree_points_Aug_18.shp")
st_crs(survey_trees_plot) <- 2253 #MI state plane (feet)

survey_trees_single <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Individual_trees_AUG_18_shape/survey.shp")
street_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Street_trees/Street_trees.shp")
arb_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Arb_trees/Arb_trees.shp")
campus_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Campus_trees/Campus_trees.shp")

# reproject layers
survey_trees_single <- st_transform(survey_trees_single, st_crs(survey_trees_plot))
street_trees <- st_transform(street_trees, st_crs(survey_trees_plot))
arb_trees <- st_transform(arb_trees, st_crs(survey_trees_plot))
campus_trees <- st_transform(campus_trees, st_crs(survey_trees_plot))

# check for projection consistency
print(st_crs(survey_trees_plot) == st_crs(survey_trees_single))
print(st_crs(survey_trees_plot) == st_crs(street_trees))
print(st_crs(survey_trees_plot) == st_crs(arb_trees))
print(st_crs(survey_trees_plot) == st_crs(campus_trees))
```

Next, we clean up and combine the survey trees layers:
```{r}
# remove unneeded columns
survey_trees_plot2 <- survey_trees_plot %>% select(c('Gns_n_S', 'DBH', 'geometry'))
survey_trees_single2 <- survey_trees_single %>% select(c('genus_and_', 'dbh', 'geometry'))

# rename remaining columns so that rbind will work
colnames(survey_trees_plot2) <- c('Project_Code', 'DBH', 'geometry')
colnames(survey_trees_single2) <- c('Project_Code', 'DBH', 'geometry')

# combine the layers into a single layer
survey_trees_full <- rbind(survey_trees_plot2, survey_trees_single2)

# remove whitespace from species code
survey_trees_full$Project_Code <- trimws(survey_trees_full$Project_Code, which = "both", whitespace = "[ \t\r\n]")

# correct some typos:
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QUBU"] <- "QUBI"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QURE"] <- "QURU"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QUPA"] <- "QUMA"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QUVA"] <- "QUVE"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QUERCUS"] <- "QUERC"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "SAAL"] <- "SAAL5"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "ULNUS"] <- "ULMUS"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "ULUMS"] <- "ULMUS"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "UPLU"] <- "ULPU"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "CAL"] <- "CALA"

# join to full tree codes
survey_trees_full <- survey_trees_full %>% left_join(tree_codes)
survey_trees_full['Source'] <- "Survey" #add source info
```

Now, we code and clean the street trees:
```{r}
# go through the "botanical" column and change any 'species' entries to 'sp.'
street_trees$BOTANICAL <- gsub('species', 'sp.', street_trees$BOTANICAL)

# join street trees to project codes
street_trees <- street_trees %>% left_join(tree_codes, by = c('BOTANICAL' = 'Botanical'))

# more cleaning
street_trees$DBH[street_trees$DBH == "N/A"] <- NA
street_trees$DBH1 <- as.numeric(str_extract(street_trees$DBH, "\\d+"))
street_trees$DBH2 <- as.numeric(str_extract(street_trees$DBH, "\\d+$"))
street_trees$DBH <- (street_trees$DBH1 + street_trees$DBH2) / 2
street_trees <- street_trees[street_trees$COMMONGENU != "Vacant",] #remove vacant tree sites
street_trees <- street_trees[street_trees$COMMONGENU != "Stump",] #remove vacant tree sites

# select only needed columns and add source info
street_trees['Source'] <- "Street Trees"
```

Code and clean the Arb trees:
```{r}
# again, sub out 'spp.' for 'sp.' so the table join works properly
arb_trees$BOTANICAL <- gsub('spp.', 'sp.', arb_trees$BOTANICAL)

# boot out dead trees
arb_trees <- arb_trees[arb_trees$COND != "Dead", ]
arb_trees <- arb_trees[arb_trees$PLANTHABIT == "Tree - canopy", ]
arb_trees <- arb_trees[arb_trees$MAINT != "Remove", ]

arb_trees <- arb_trees %>% left_join(tree_codes, by = c('BOTANICAL' = 'Botanical'))
arb_trees['Source'] <- "Arboretum"
```

Code and clean the campus trees:
```{r}
# remove NA and unknown and removed trees
campus_trees <- campus_trees[!is.na(campus_trees$d_SpeciesI), ]
campus_trees <- campus_trees[campus_trees$d_SpeciesI != 'Unknown', ]
campus_trees <- campus_trees[campus_trees$d_Removed == "No or False", ]

# make data easier to work with
campus_trees <- campus_trees %>% select('d_SpeciesI', 'Diameter', 'geometry')
colnames(campus_trees) <- c('Common_Name', 'DBH', 'geometry')

# rename a bunch of trees to make the table join possible
campus_trees$Common_Name[campus_trees$Common_Name == "Colorado Blue Spruce"] <- "Blue Spruce"
campus_trees$Common_Name[campus_trees$Common_Name == "Honeylocust (Thornless)"] <- "Honeylocust, Thornless"
campus_trees$Common_Name[campus_trees$Common_Name == "Honey Locust"] <- "Honeylocust, Thornless"
campus_trees$Common_Name[campus_trees$Common_Name == "Crabapple"] <- "Crabapple, Common"
campus_trees$Common_Name[campus_trees$Common_Name == "Mary Potter Crabapple"] <- "Crabapple, Common"
campus_trees$Common_Name[campus_trees$Common_Name == "Northern red oak"] <- "Northern Red Oak"
campus_trees$Common_Name[campus_trees$Common_Name == "White Pine"] <- "Eastern White Pine"
campus_trees$Common_Name[campus_trees$Common_Name == "Littleleaf Linden"] <- "Linden, Littleleaf"
campus_trees$Common_Name[campus_trees$Common_Name == "Japanese Cherry"] <- "Plum" #to classify to prunus sp.
campus_trees$Common_Name[campus_trees$Common_Name == "Cherry Hybrid"] <- "Plum" #ditto
campus_trees$Common_Name[campus_trees$Common_Name == "American Arborvitae"] <- "Northern White-Cedar"
campus_trees$Common_Name[campus_trees$Common_Name == "London Planetree"] <- "Planetree, London"
campus_trees$Common_Name[campus_trees$Common_Name == "Heritage River Birch"] <- "River Birch"
campus_trees$Common_Name[campus_trees$Common_Name == "Black Gum"] <- "Blackgum"
campus_trees$Common_Name[campus_trees$Common_Name == "Serbian Spruce"] <- "Spruce, Serbian"
campus_trees$Common_Name[campus_trees$Common_Name == "Shadblow Serviceberry"] <- "Serviceberry"
campus_trees$Common_Name[campus_trees$Common_Name == "Saucer Magnolia"] <- "Magnolia, Chinese; Saucer"
campus_trees$Common_Name[campus_trees$Common_Name == "Ginkgo"] <- "Ginkgo, Maidenhair Tree"
campus_trees$Common_Name[campus_trees$Common_Name == "Tuliptree"] <- "Yellow-Poplar"
campus_trees$Common_Name[campus_trees$Common_Name == "Bradford Pear"] <- "Pear, Callery"
campus_trees$Common_Name[campus_trees$Common_Name == "American Arborvitae \"Nigra\""] <- "northern white-cedar"
campus_trees$Common_Name[campus_trees$Common_Name == "Red Sunset Maple"] <- "Maple"
campus_trees$Common_Name[campus_trees$Common_Name == "Whitespire Birch"] <- "birch sp."
campus_trees$Common_Name[campus_trees$Common_Name == "Callery Pear"] <- "Pear, Callery"
campus_trees$Common_Name[campus_trees$Common_Name == "Common Hackberry"] <- "hackberry"
campus_trees$Common_Name[campus_trees$Common_Name == "Japanese Zelkova
"] <- "zelkova, Japanese"
campus_trees$Common_Name[campus_trees$Common_Name == "Cleveland Select Pear"] <- "Pear"
campus_trees$Common_Name[campus_trees$Common_Name == "Crusader Hawthorn"] <- "Hawthorn"
campus_trees$Common_Name[campus_trees$Common_Name == "Japanese Tree Lilac"] <- "lilac, Japanese tree"
campus_trees$Common_Name[campus_trees$Common_Name == "American Liberty Elm"] <- "American elm"
campus_trees$Common_Name[campus_trees$Common_Name == "Amur Maple"] <- "Maple, Amur"

campus_trees <- campus_trees %>% left_join(tree_codes)
campus_trees['Source'] <- "Campus"
```

Here we do a few checks to make sure the data cleaning went well:
```{r}
ctna <- campus_trees[is.na(campus_trees$Plant_Code), ] %>%
  group_by(Common_Name) %>%
  summarise(n = n())

survey_counts <- survey_trees_full %>%
  group_by(Plant_Code) %>%
  summarise(n = n())
```

Choose final variables to keep and combine into a single master layer:
```{r}
survey_trees_full <- survey_trees_full %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))
street_trees <- street_trees %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))
arb_trees <- arb_trees %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))
campus_trees <- campus_trees %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))

training_trees_master <- rbind(survey_trees_full, arb_trees, street_trees, campus_trees)
# st_write(training_trees_master, "C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/training_trees_master.shp")
```
Now we reload the file (to avoid having to repeat previous steps) and generate some graphs and charts of our data:
```{r}
training_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/training_trees_master.shp")
```

Here we get counts for the top 20 trees in our full dataset:
```{r}
tree_counts_full <- training_trees %>%
  group_by(Cmmn_Nm) %>%
  summarize(tree_count = n(), mean_dbh = mean(DBH, na.rm = TRUE)) 

tree_counts_full <- arrange(tree_counts_full, desc(tree_count))
tree_counts_full_top20 <- tree_counts_full[1:20,]

ggplot(tree_counts_full_top20, aes(x=fct_reorder(Cmmn_Nm, tree_count), y=tree_count)) + geom_bar(stat = 'identity', fill="navy") + coord_flip() + labs(x="Tree Species", y="Tree Count", title="Top 20 Training Data Trees")
```
And here we get counts for the top 20 trees in our survey:
```{r}
survey_counts <- training_trees %>%
  filter(Source == 'Survey') %>%
  group_by(Cmmn_Nm) %>%
  summarize(tree_count = n(), mean_dbh = mean(DBH, na.rm = TRUE)) 

survey_counts <- arrange(survey_counts, desc(tree_count))
survey_counts_top20 <- survey_counts[1:20,]

ggplot(survey_counts_top20, aes(x=fct_reorder(Cmmn_Nm, tree_count), y=tree_count)) + geom_bar(stat = 'identity', fill="forestgreen") + coord_flip() + labs(x="Tree Species", y="Tree Count", title="Top 20 Trees Surveyed")
```

```{r}
ggplot(training_trees[training_trees$Genus == 'QUERC',], aes(x=DBH)) + geom_histogram(fill="goldenrod", colour = 'black', binwidth = 5) + labs(title = "Oak DBH distribution")
```
```{r}
ggplot(training_trees[training_trees$Genus == 'ACER',], aes(x=DBH)) + geom_histogram(fill="coral4", colour = 'black', binwidth = 5) + labs(title = "Maple DBH distribution")
```


```{r}
ggplot(training_trees[training_trees$Genus == 'CARYA',], aes(x=DBH)) + geom_histogram(fill="forestgreen", colour = 'black', binwidth = 5) + labs(title = "Hickory DBH distribution")
```



































