---
title: "Master Training Data Maker"
author: "Thomas Estabrook"
date: "8/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(stringr)
library(sf)
library(cartography)
library(tidyverse)
```

Goal: Get every dataset into a single, usable shapefile. This means that all trees have their proper code, invalid or dead trees are eliminated. The final file should have the following fields:
- Tree code
- DBH
- Source file
- Geometry

Outline:
1. Load in survey trees shapefile (output of survey_to_points.rmd), campus trees, arb trees, and street trees
  a. Get all trees into the correct coordinate reference system (crs)
2. Survey trees
  a. Combine with single point layer
  b. Eliminate typos
3. Campus trees, Arb trees, Street trees
  a. Recode to project tree codes
  b. Calculate proper DBH (campus)
  c. Eliminate extraneous columns
  d. Eliminate dead trees
4. Merge all into single layer

Notes
- NEED TO FILTER OVERTOPPED TREES FROM SURVEY DATA
- The table joins will produce duplicate entries UNLESS the key columns include only unique values. Unfortunately, for these datasets we will be using the `Project_Code`, `Botanical`, and `Common_Name` columns for joins, so they all need to only have unique entries. If using this script, make sure you have the most recent version of `GenusSpecies_PlantCode_TGE_edits.csv`. 

Join Keys:
- Survey: Project_Code
- Arb: Botanical
- Street trees: Botanical
- Campus trees: Campus_Name
----------------------------------------------------------------------

First we load in the necessary training data layers and ensure they are all in the same projection:
```{r}
# set up tree codes table
tree_codes <- read.csv("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data//CodedTrainingData/Tree_codes_master_09-06.csv")
# colnames(tree_codes)[1] <- "Common_Name"
# tree_codes$Project_Code[tree_codes$Project_Code == ""] <- NA
# tree_codes$Plant_Code[tree_codes$Plant_Code == ""] <- NA
# tree_codes <- tree_codes %>%
#   mutate(Project_Code = coalesce(Project_Code, Plant_Code)) %>%
#   mutate(Common_Name = str_to_title(Common_Name))

# For the joins to work, we need to check for duplicates in the project codes. If any of these tables have an entry higher than one, you will need to adjust the plant code document accordingly. 
n_occur_code <- data.frame(table(tree_codes$Project_Code))
n_occur_common <- data.frame(table(tree_codes$Common_Name))
n_occur_botanical <- data.frame(table(tree_codes$Botanical))
max(n_occur_code$Freq)
max(n_occur_common$Freq)
max(n_occur_botanical$Freq)  # If all are one, we're good

# load data layers
survey_trees_plot <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/tree_points_Aug_18.shp")
st_crs(survey_trees_plot) <- 2253 #MI state plane (feet)

survey_trees_single <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Individual_trees_AUG_18_shape/survey.shp")
street_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Street_trees/Street_trees.shp")
arb_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Arb_trees/Arb_trees.shp")
campus_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Campus_trees/Campus_trees.shp")
radrick_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Radrick_stemmap_Project.geojson")

# reproject layers
survey_trees_single <- st_transform(survey_trees_single, st_crs(survey_trees_plot))
street_trees <- st_transform(street_trees, st_crs(survey_trees_plot))
arb_trees <- st_transform(arb_trees, st_crs(survey_trees_plot))
campus_trees <- st_transform(campus_trees, st_crs(survey_trees_plot))
radrick_trees <- st_transform(radrick_trees, st_crs(survey_trees_plot))

# check for projection consistency
print(st_crs(survey_trees_plot) == st_crs(survey_trees_single))
print(st_crs(survey_trees_plot) == st_crs(street_trees))
print(st_crs(survey_trees_plot) == st_crs(arb_trees))
print(st_crs(survey_trees_plot) == st_crs(campus_trees))
```

Next, we clean up and combine the survey trees layers:
```{r}
# remove unneeded columns
survey_trees_plot2 <- survey_trees_plot %>% select(c('Gns_n_S', 'DBH', 'geometry'))
survey_trees_single2 <- survey_trees_single %>% select(c('genus_and_', 'dbh', 'geometry'))

# rename remaining columns so that rbind will work
colnames(survey_trees_plot2) <- c('Project_Code', 'DBH', 'geometry')
colnames(survey_trees_single2) <- c('Project_Code', 'DBH', 'geometry')

# combine the layers into a single layer
survey_trees_full <- rbind(survey_trees_plot2, survey_trees_single2)

# remove whitespace from species code
survey_trees_full$Project_Code <- trimws(survey_trees_full$Project_Code, which = "both", whitespace = "[ \t\r\n]")

# correct some typos:
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QUBU"] <- "QUBI"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QURE"] <- "QURU"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QUPA"] <- "QUMA"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QUVA"] <- "QUVE"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "QUERCUS"] <- "QUERC"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "SAAL"] <- "SAAL5"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "ULNUS"] <- "ULMUS"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "ULUMS"] <- "ULMUS"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "UPLU"] <- "ULPU"
survey_trees_full$Project_Code[survey_trees_full$Project_Code == "CAL"] <- "CALA"

# join to full tree codes
survey_trees_full <- survey_trees_full %>% left_join(tree_codes)
survey_trees_full['Source'] <- "Survey" #add source info
```

Now, we code and clean the street trees:
```{r}
# go through the "botanical" column and change any 'species' entries to 'sp.'
street_trees$BOTANICAL <- gsub('species', 'sp.', street_trees$BOTANICAL)

# join street trees to project codes
street_trees <- street_trees %>% left_join(tree_codes, by = c('BOTANICAL' = 'Botanical'))

# more cleaning
street_trees$DBH[street_trees$DBH == "N/A"] <- NA
street_trees$DBH1 <- as.numeric(str_extract(street_trees$DBH, "\\d+"))
street_trees$DBH2 <- as.numeric(str_extract(street_trees$DBH, "\\d+$"))
street_trees$DBH <- (street_trees$DBH1 + street_trees$DBH2) / 2
street_trees <- street_trees[street_trees$COMMONGENU != "Vacant",] #remove vacant tree sites
street_trees <- street_trees[street_trees$COMMONGENU != "Stump",] #remove vacant tree sites

# select only needed columns and add source info
street_trees['Source'] <- "Street Trees"
```

Code and clean the Arb trees:
```{r}
# again, sub out 'spp.' for 'sp.' so the table join works properly
arb_trees$BOTANICAL <- gsub('spp.', 'sp.', arb_trees$BOTANICAL)

# boot out dead trees
arb_trees <- arb_trees[arb_trees$COND != "Dead", ]
arb_trees <- arb_trees[arb_trees$PLANTHABIT == "Tree - canopy", ]
arb_trees <- arb_trees[arb_trees$MAINT != "Remove", ]

arb_trees <- arb_trees %>% left_join(tree_codes, by = c('BOTANICAL' = 'Botanical'))
arb_trees['Source'] <- "Arboretum"
```

Code and clean the campus trees:
```{r}
# remove NA and unknown and removed trees
campus_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Campus_trees/Campus_trees.shp")

campus_key <- read.csv("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data//CodedTrainingData/campus_join_key.csv")
campus_key$Campus_Name <- trimws(campus_key$Campus_Name, which = "both", whitespace = "[ \t\r\n]")

campus_trees <- campus_trees[!is.na(campus_trees$d_SpeciesI), ]
campus_trees <- campus_trees[campus_trees$d_SpeciesI != 'Unknown', ]
campus_trees <- campus_trees[campus_trees$d_Removed == "No or False", ]

# make data easier to work with
campus_trees <- campus_trees %>% select('d_SpeciesI', 'Diameter', 'geometry')
colnames(campus_trees) <- c('Campus_Name', 'DBH', 'geometry')
campus_trees$Campus_Name <- str_to_title(campus_trees$Campus_Name)

campus_trees <- campus_trees %>%
  left_join(campus_key) %>%
  select("Common_Name", 'DBH', 'geometry')

campus_trees <- campus_trees %>% left_join(tree_codes, by = "Common_Name")
campus_trees['Source'] <- "Campus"

ctna <- campus_trees[is.na(campus_trees$Genus), ] %>%
  group_by(Genus) %>%
  summarise(n = n())
```

Code and clean Radrick trees:
```{r}
radrick_trees <- radrick_trees %>%
  select('Species', 'Diameter', 'geometry') %>%
  rename(Project_Code = Species, DBH = Diameter) %>%
  left_join(tree_codes) %>%
  mutate(Source = "Radrick")

```


Choose final variables to keep and combine into a single master layer:
```{r}
survey_trees_full <- survey_trees_full %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))
street_trees <- street_trees %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))
arb_trees <- arb_trees %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))
campus_trees <- campus_trees %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))
radrick_trees <- radrick_trees %>% select(c('Plant_Code', 'Common_Name', 'DBH', 'Source', 'Genus', 'Project_Code', 'geometry'))

training_trees_master <- rbind(survey_trees_full, arb_trees, street_trees, campus_trees, radrick_trees)
st_write(training_trees_master, "C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/training_trees_master_10-6v2.shp")
```


------------------------------------------------------------------------------------------------------------------------


Now we reload the file (to avoid having to repeat previous steps) and generate some graphs and charts of our data:
```{r}
training_trees <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/training_trees_master.shp")
#training_trees <- read_sf("C:/Users/Thomas/OneDrive - Umich/School/Capstone Project/Data/training_trees_master.shp")
```

Here we get counts for the top 20 trees in our full dataset:
```{r}
tree_counts_full <- training_trees %>%
  group_by(Cmmn_Nm) %>%
  summarize(tree_count = n(), mean_dbh = mean(DBH, na.rm = TRUE)) 

tree_counts_full <- arrange(tree_counts_full, desc(tree_count))
tree_counts_full_top20 <- tree_counts_full[1:20,]

ggplot(tree_counts_full_top20, aes(x=fct_reorder(Cmmn_Nm, tree_count), y=tree_count)) + geom_bar(stat = 'identity', fill="navy") + coord_flip() + labs(x="Tree Species", y="Tree Count", title="Top 20 Training Data Trees")
```
And here we get counts for the top 20 trees in our survey:
```{r}
survey_counts <- training_trees %>%
  filter(Source == 'Survey') %>%
  group_by(Cmmn_Nm) %>%
  summarize(tree_count = n(), mean_dbh = mean(DBH, na.rm = TRUE)) 

survey_counts <- arrange(survey_counts, desc(tree_count))
survey_counts_top20 <- survey_counts[1:20,]

ggplot(survey_counts_top20, aes(x=fct_reorder(Cmmn_Nm, tree_count), y=tree_count)) + geom_bar(stat = 'identity', fill="forestgreen") + coord_flip() + labs(x="Tree Species", y="Tree Count", title="Top 20 Trees Surveyed")
```

```{r}
ggplot(training_trees[training_trees$Genus == 'QUERC',], aes(x=DBH)) + geom_histogram(fill="goldenrod", colour = 'black', binwidth = 5) + labs(title = "Oak DBH distribution")
```
```{r}
ggplot(training_trees[training_trees$Genus == 'ACER',], aes(x=DBH)) + geom_histogram(fill="coral4", colour = 'black', binwidth = 5) + labs(title = "Maple DBH distribution")
```


```{r}
ggplot(training_trees[training_trees$Genus == 'CARYA',], aes(x=DBH)) + geom_histogram(fill="forestgreen", colour = 'black', binwidth = 5) + labs(title = "Hickory DBH distribution")
```

```{r}
ggplot(training_trees[(training_trees$Genus == 'QUERC') & (training_trees$Source == 'Survey'),], aes(x=DBH)) + geom_histogram(fill="goldenrod", colour = 'black', binwidth = 5) + labs(title = "Oak DBH distribution (Survey)")
```


```{r}
ggplot(training_trees[(training_trees$Genus == 'QUERC') & (training_trees$Source == 'Street Trees'),], aes(x=DBH)) + geom_histogram(fill="goldenrod", colour = 'black', binwidth = 5) + labs(title = "Oak DBH distribution (Street Trees)")
```

```{r}
ggplot(training_trees[(training_trees$Genus == 'QUERC'),], aes(x=DBH, color=Source, fill = Source)) + geom_histogram(binwidth = 5, position = "identity", alpha = 0.25) + labs(title = "Oak DBH distribution")
```

```{r}
ggplot(drop_na(training_trees[(training_trees$Genus == 'QUERC'),], Source), aes(x=DBH, fill = Source)) + geom_histogram(binwidth = 5, position = "identity", alpha = 0.5, color="black") + facet_wrap(Source ~ ., ncol = 2, scales='free_y') + labs(title = "Oak DBH distribution") + scale_fill_brewer(palette = "Set1")
```

```{r}
ggplot(drop_na(training_trees[(training_trees$Genus == 'CARYA'),], Source), aes(x=DBH, fill = Source)) + geom_histogram(binwidth = 5, position = "identity", alpha = 0.5, color="black") + facet_wrap(Source ~ ., ncol = 2, scales='free_y') + labs(title = "Hickory DBH distribution") + scale_fill_brewer(palette = "Set2")
```

```{r}
ggplot(drop_na(training_trees[(training_trees$Genus == 'ACER'),], Source), aes(x=DBH, fill = Source)) + geom_histogram(binwidth = 5, position = "identity", alpha = 0.5, color="black") + facet_wrap(Source ~ ., ncol = 2, scales='free_y') + labs(title = "Maple DBH distribution") + scale_fill_brewer(palette = "Set3")
```

```{r}
ggplot(drop_na(training_trees[(training_trees$Genus %in% c('ACER', 'QUERC', 'CARYA')),], Source), aes(x=DBH, fill = Source)) + geom_histogram(binwidth = 5, position = "identity", alpha = 0.4, color="black") + facet_grid(Source ~ Genus, scales='free_y') + labs(title = "DBH Distribution by Data Source", y = "Count") + scale_fill_brewer(palette = "Set1") + theme(legend.position = 'none')

ggsave(filename = "C:/Users/User/OneDrive - Umich/School/Capstone Project/grid_plot.png", width = 2400, height = 1600, units = 'px', dpi = 300, bg = "#384f37")
```

```{r}
ggplot(drop_na(training_trees[(training_trees$Genus %in% c('QUERC')) & (training_trees$Source %in% "Survey"),], Source), aes(x=DBH, fill = Plnt_Cd)) + geom_histogram(binwidth = 5, position = "identity", alpha = 0.4, color="black") + facet_grid(Plnt_Cd ~ .) + labs(title = "DBH Distribution by Data Source", y = "Count") + scale_fill_brewer(palette = "Set1") + theme(legend.position = 'none')
```
```{r}
boxplot(DBH ~ Plnt_Cd, data = drop_na(training_trees[(training_trees$Genus %in% c('QUERC')),], Source))
```



















