---
title: "Tree Survey to Point Layer"
author: "Thomas Estabrook"
date: "7/7/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sf)
library(dplyr)
library(cartography)
library(ggplot2)
```

Outline:
- Write function for getting coordinates from sample plot location and x and y

- Create polygons with plot outlines
  - Get locations for each corner and store in variables
    - 100 ft north
    - 100 ft east
    - 100 ft north and east
  - Create polygon from coordinates
  
- Place points for each tree
  - Loop through each sample site
    - Subset points for that site
    - Loop through each tree in the site
      - Add a new point for the tree with appropriate transformation
      
      
```{r}
coord_convert <- function(old_x, old_y, plot_x, plot_y, coord){
  # declination is 7 degrees three minutes to the West = 7.05 degreees
  # assuming same units since input layer is in feet
  theta = -0.123 # convert to radians, negative for clockwise correction
  new_x <- (old_x*cos(theta)) - (old_y*sin(theta))
  new_y <- (old_y*cos(theta)) + (old_x*sin(theta))
  
  return(ifelse(coord == "X", plot_x + new_x, plot_y + new_y))
}
```


```{r}
# read in survey tree csv file
survey_trees <- read.csv("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/July_27_Tree_Survey/survey_0.csv")
# read in shapefile for plot locations
sample_plots <- read_sf("C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/Sample_Plots_shapefile/SamplePlots.shp")
#source CRS: EPSG 2253, units are feet
```

```{r}
#get the x and y coordinate values from the sample plot shapefile
sample_plots <- sample_plots %>%
  mutate(X_plot = st_coordinates(sample_plots[,7])[,"X"]) %>%
  mutate(Y_plot = st_coordinates(sample_plots[,7])[,"Y"]) %>%
  select(!c(Coordinate, GlobalID, CreationDa, Creator, EditDate, Editor, H_precisio, X, Y))

sample_plots <- st_drop_geometry(sample_plots)
```

```{r}
theta = 0.123 # convert to radians, negative for clockwise correction
survey_trees <- survey_trees %>%
  inner_join(sample_plots, by = c("Plot.ID" = "CID")) %>%
  select(!c(x,y,Cartesian.or.Polar.Coordinates.,GlobalID, CreationDate, Creator, EditDate, Editor,Surveyor,Notes.x,Notes.y)) %>%
  mutate(new_x = (X.value*cos(theta)) - (Y.value*sin(theta))) %>%
  mutate(new_y = (Y.value*cos(theta)) + (X.value*sin(theta))) %>%
  mutate(final_x = X_plot + new_x) %>%
  mutate(final_y = Y_plot + new_y)

```


```{r}
proj_crs <- st_crs(sample_plots)
tree_points <- st_as_sf(survey_trees, coords = c("final_x", "final_y"), crs=proj_crs)
```

```{r}
ggplot() +
  geom_sf(data = tree_points) +
  ggtitle("Map of Plot Locations")
```

```{r}
write_sf(tree_points, "C:/Users/User/OneDrive - Umich/School/Capstone Project/Data/tree_points_test.shp")
```









